```{r}
#install.packages('readxl')
library('readxl')
data <- readxl::read_xlsx('GSS.xlsx')
```

```{r}
library('dplyr')
library('tidyverse')
library('ggplot2')
```

```{r}
str(data)

data <- rename(data, 'Education'='Highest year of school completed')
data <- rename(data, 'Year'='Gss year for this respondent')
data <- rename(data, 'Income'='Total family income')
data <- rename(data, 'Race' = 'Race of respondent')
data <- rename(data, 'Sex' = 'Respondents sex')
data <- rename(data, 'Party' = 'Political party affiliation')
data <- rename(data, 'Happiness' = 'General happiness')
data <- rename(data, 'Marital.status' = 'Marital status')
data <- rename(data, 'ID' = 'Respondent id number')
data <- rename(data, 'Ballot' = 'Ballot used for interview')

str(data)
```
### Question 3
##### How is the happiness of a respondent related to his/her marriage status? Define that a person is happy if the response to question is “Very happy” or “Pretty happy”.
```{r}
df <- as.data.frame(data)

happy_df <- df %>% 
  group_by(Marital.status) %>%
  filter(Happiness=="Pretty happy" | Happiness=="Very happy") %>%
  filter(Marital.status!="No answer")

unhappy_df <- df %>%
  group_by(Marital.status) %>%
  filter(Happiness=="Not too happy") %>%
  filter(Marital.status!="No answer")

unhappy_table <- unhappy_df %>%
  group_by(Marital.status) %>%
  summarize(
    Total_Unhappy=n()
  )
unhappy_table

happy_table <- happy_df %>%
  group_by(Marital.status) %>%
  summarize(
    Total_Happy=n()
  )
happy_table

ggplot(happy_df, aes(x=Marital.status)) + geom_bar(color="blue", fill="red")
ggplot(unhappy_df, aes(x=Marital.status)) + geom_bar(color="blue", fill="red")

mosaicplot(happy_df$Happiness~happy_df$Marital.status, shade=TRUE, las=1)
mosaicplot(unhappy_df$Happiness~unhappy_df$Marital.status, shade=TRUE, las=1)
```
<h3>Conclusion</h3>
<p>
It would seem that there are more people married and happy than there are of any other group. Never married people also seem much happier than divorced, separated, or divorced people. </p>

### Question 4
##### Does party affiliation affect the relationship you found in the last question?
```{r}
unique(happy_df$Party)

happy_indep_df <- happy_df %>%
  group_by(Marital.status) %>%
  filter(Party=="Independent" | Party=="Ind,near rep" | Party=="Ind,near dem")
happy_indep_df
ti1 <- happy_indep_df %>%
  summarize(
    Total_Happy_Indep=n()
  )
ti1

unhappy_indep_df <- unhappy_df %>%
  group_by(Marital.status) %>%
  filter(Party=="Independent" | Party=="Ind,near rep" | Party=="Ind,near dem")
unhappy_indep_df
ti2 <- unhappy_indep_df %>%
  summarize(
    Total_Unhappy_Indep=n()
  )
ti2

happy_rep_df <- happy_df %>%
  group_by(Marital.status) %>%
  filter(Party=="Strong republican" | Party=="Not str republican")
t1 <- happy_rep_df %>%
  summarize(
    Total_Happy_Rep=n()
  )
t1

unhappy_rep_df <- unhappy_df %>%
  group_by(Marital.status) %>%
  filter(Party=="Strong republican" | Party=="Not str republican")
t2 <- unhappy_rep_df %>%
  summarize(
    Total_Unhappy_Rep=n()
  )
t2

happy_dem_df <- happy_df %>%
  group_by(Marital.status) %>%
  filter(Party=="Strong democrat" | Party=="Not str democrat")
t3 <- happy_dem_df %>%
  summarize(
    Total_Happy_Dem=n()
  )
t3

unhappy_dem_df <- unhappy_df %>%
  group_by(Marital.status) %>%
  filter(Party=="Strong democrat" | Party=="Not str democrat")
t4 <- unhappy_dem_df%>%
  summarize(
    Total_Unhappy_Dem=n()
  )
t4

ggplot(happy_rep_df, aes(x=Party)) + geom_bar(color="blue", fill="red")
ggplot(unhappy_rep_df, aes(x=Party)) + geom_bar(color="blue", fill="red")
ggplot(happy_dem_df, aes(x=Party)) + geom_bar(color="red", fill="blue")
ggplot(unhappy_dem_df, aes(x=Party)) + geom_bar(color="red", fill="blue")
ggplot(happy_indep_df, aes(x=Party)) + geom_bar(color="black", fill="purple")
ggplot(unhappy_indep_df, aes(x=Party)) + geom_bar(color="black", fill="purple")
```
<h3>Conclusion</h3>
<p>
It would seem that there are similar amounts of happy and unhappy democrats as well as happy and unhappy republicans as well as happy and unhappy independents. This would seem to suggest party affiliation is not a factor in happiness.</p>
