---
title: "Lab 3"
author: "Brian Sayre, Desmond Orton, Caitlin Fordell"
date: "3/5/2021"
output: html_document
---


## Setup
```{r}
library(dplyr)
library(tidyverse)
library(ggplot2)
data <- readxl::read_xls('./GSS.xls')
str(data)
df <- data.frame(
  year=data$`Gss year for this respondent`, 
  happy=data$`General happiness`, 
  party=data$`Political party affiliation`, 
  income=data$`Total family income`, 
  race=data$`Race of respondent`, 
  sex=data$`Respondents sex`, 
  school=data$`Highest year of school completed`, 
  mstatus=data$`Marital status`, 
  id=data$`Respondent id number`, 
  ballot=data$`Ballot used for interview`) 
str(df)
```


### Question 1
Average years of education for respondents with each marital status.

```{r}
q1df <- as.data.frame(df)
q1df <- filter(q1df, school != 'No answer' & school != "Don't know" & !is.na(school) & school != "NA" )
unique(q1df$school)
q1df$school <- as.numeric(q1df$school)
ggplot(q1df, aes(y=school, x=mstatus), fill=school) + geom_boxplot()+xlab("Marital Status")+ylab("Education")+ggtitle("Marital Status vs. Education")
```

```{r}
q1df %>% 
  group_by(mstatus) %>% 
  summarize(
    mean_rate = mean(school, na.rm=TRUE),
    sd_rate = sd(school, na.rm=TRUE),
    nrows = n()
  )
```

I chose to display the data as a boxplot for each marriage status. I chose this to show as mush information about the education for each status as possible. We can see that the ones with no answer seem to have the highest mean education which is something I did not expect. We also can see the rest have around the same mean, but a difference in most of their interquartile ranges. Divorced, married and never married seem to have a higher IQR than separated and widowed.

### Question 2
Creating a data frame for better educated respondents for each marital status
```{r}
q2df <- q1df %>% 
  group_by(mstatus) %>% 
  filter(school > mean(school))
str(q2df)
```


### Question 3
How is happiness related to marital status?
```{r}
df2 <- as.data.frame(df)
str(df2)
happy_df <- df2 %>% 
  group_by(mstatus) %>%
  filter(happy=="Pretty happy" | happy=="Very happy") %>%
  filter(mstatus!="No answer")

unhappy_df <- df2 %>%
  group_by(mstatus) %>%
  filter(happy=="Not too happy") %>%
  filter(mstatus!="No answer")

unhappy_table <- unhappy_df %>%
  group_by(mstatus) %>%
  summarize(
    Total_Unhappy=n()
  )
unhappy_table

happy_table <- happy_df %>%
  group_by(mstatus) %>%
  summarize(
    Total_Happy=n()
  )
happy_table

ggplot(happy_df, aes(x=mstatus)) + geom_bar(color="blue", fill="red") + ggtitle("Happiness by Marital Status")
ggplot(unhappy_df, aes(x=mstatus)) + geom_bar(color="blue", fill="red")  + ggtitle("UnHappiness by Marital Status")

mosaicplot(happy_df$happy~happy_df$mstatus, shade=TRUE, las=1, main="Mosaic of Happy to Marital Status")
mosaicplot(unhappy_df$happy~unhappy_df$mstatus, shade=TRUE, las=1, main="Mosaic of UnHappy to Marital Status")
```
<p>
It would seem that there are more people married and happy than there are of any other group. Never married people also seem much happier than divorced, separated, or divorced people. </p>

### Question 4
Does party affiliation affect the relationship between happiness and marital status?
```{r}
unique(happy_df$party)

happy_indep_df <- happy_df %>%
  group_by(mstatus) %>%
  filter(party=="Independent" | party=="Ind,near rep" | party=="Ind,near dem")
happy_indep_df
ti1 <- happy_indep_df %>%
  summarize(
    Total_Happy_Indep=n()
  )
ti1

unhappy_indep_df <- unhappy_df %>%
  group_by(mstatus) %>%
  filter(party=="Independent" | party=="Ind,near rep" | party=="Ind,near dem")
unhappy_indep_df
ti2 <- unhappy_indep_df %>%
  summarize(
    Total_Unhappy_Indep=n()
  )
ti2

happy_rep_df <- happy_df %>%
  group_by(mstatus) %>%
  filter(party=="Strong republican" | party=="Not str republican")
t1 <- happy_rep_df %>%
  summarize(
    Total_Happy_Rep=n()
  )
t1

unhappy_rep_df <- unhappy_df %>%
  group_by(mstatus) %>%
  filter(party=="Strong republican" | party=="Not str republican")
t2 <- unhappy_rep_df %>%
  summarize(
    Total_Unhappy_Rep=n()
  )
t2

happy_dem_df <- happy_df %>%
  group_by(mstatus) %>%
  filter(party=="Strong democrat" | party=="Not str democrat")
t3 <- happy_dem_df %>%
  summarize(
    Total_Happy_Rep=n()
  )
t3

unhappy_dem_df <- unhappy_df %>%
  group_by(mstatus) %>%
  filter(party=="Strong democrat" | party=="Not str democrat")
t4 <- unhappy_dem_df%>%
  summarize(
    Total_Unhappy_Rep=n()
  )
t4

ggplot(happy_rep_df, aes(x=party)) + geom_bar(color="blue", fill="red") + ggtitle("Happy Republicans")
ggplot(unhappy_rep_df, aes(x=party)) + geom_bar(color="blue", fill="red") + ggtitle("Unhappy Republicans")
ggplot(happy_dem_df, aes(x=party)) + geom_bar(color="red", fill="blue") + ggtitle("Happy Democrats")
ggplot(unhappy_dem_df, aes(x=party)) + geom_bar(color="red", fill="blue") + ggtitle("Unhappy Democrats")
ggplot(happy_indep_df, aes(x=party)) + geom_bar(color="black", fill="purple") + ggtitle("Happy Independents")
ggplot(unhappy_indep_df, aes(x=party)) + geom_bar(color="black", fill="purple") + ggtitle("Unhappy Independents")
```

It would seem that there are similar amounts of happy and unhappy democrats as well as happy and unhappy republicans as well as happy and unhappy independents. This would seem to suggest party affiliation is not a factor in happy.

## Question 5
What is the relationship between martial status and education?
```{r}
##creating data frame for Q5
q5df <- as.data.frame(df)
q5df <- filter(q5df, school != 'No answer' & school != "Don't know" & !is.na(school) & school != "NA" )
unique(q5df$school)
q5df$school <- as.numeric(q5df$school)
```



```{r}
##plot of marital status and education
ggplot(data=q5df, aes(x=mstatus,y=school))+geom_boxplot()+ggtitle("Marital Status vs. Education")+theme(axis.text.x = element_text(angle = 90))+xlab("Marital Status")+ylab("Education")
```

The group with the highest median years of education was the "No Answer" group, while the group with the lowest median years of education was Widowed people. Since the no answer group likely has less respondents than the other groups, they are likely to get more average responses for years of schooling than the other groups.

```{r}
ggplot(data=q5df, aes(x=mstatus, y=school))+geom_bar(stat = 'identity')+xlab("Marital Status")+ggtitle("Marital Status vs. Total Education Bar Graph")+ylab("Education")
```

Overall, the group of married people has the highest total amount of education, distantly followed by the never married group. Since this is the total education though, we can not tell if this is solely due to married people having more education or if there were just many more married people than other groups. This graph also shows that the total education of the No Answer group is much less than the total education for all other groups.

## Question 6A
What is the relationship between race and education?
```{r}
##exploring relationship between race and education
q6dfA <- select(df, race, school)
q6dfA <- filter(q6dfA, school != 'No answer' & school != "Don't know" & !is.na(school) & school != "NA" )
```

```{r}
##exploring school variable
q6dfA$school <- as.numeric(q6dfA$school)
ggplot(q6dfA, aes(x=school))+geom_bar()+xlab("Education")+ ylab("Count")+ggtitle("Counts of Years of Education")
```

The most common amount of education of respondents for the study was 12 years of school, which means that more respondents received a high school education than any other number of years of education. The data is skewed to the left with a center around 12 years of education.

```{r}
##exploring race variable
ggplot(q6dfA, aes(x=race))+geom_bar()+xlab("Race")+ylab("Count")+ggtitle("Number of Respondents per Race")+xlab("Race")+ylab("Education")
```

There are approximately 5 times as many white respondents as the next largest category (black respondents). The majority of the data collected was from white-identifying people. There is still a sizeable amount of data for both the black and other people.

```{r}
##exploring the relationship between race and school
ggplot(data=q6dfA, aes(x=race, y=school))+geom_boxplot()+ggtitle("Race vs. Education")
```

The median years of education for both black and white people is about 12 years while the median years of education for the other races is 13 years. Overall, black people tended to have lower education than white people and the other races had the widest range. This could be because the other category includes other races like latino and asian. 

### Question 6b
Does income affect your overall happiness?

```{r}
q6df <- select(df,happy, income)
q6df <- filter(q6df, happy != "No answer" & happy != "Don't know" & happy != "Not applicable")
q6df <- filter(q6df, income != "No answer" & income != "Don't know" & income != "Not applicable" & income != "Refused")
q6df$happyNum <- recode(q6df$happy, "Not too happy" = 1, "Pretty happy" = 2, "Very happy" = 3)
q6df <- aggregate(q6df$happyNum, by=list(q6df$income), FUN=mean)
q6qf <- rename(q6df, Income = Group.1, happy = x)
ordered <- c("Lt $1000", "$1000 to 2999", "$3000 to 3999", "$4000 to 4999", "$5000 to 5999", "$6000 to 6999", "$7000 to 7999", "$8000 to 9999", "$10000 - 14999", "$15000 - 19999", "$20000 - 24999", "$25000 or more")
q6qf$Income <- factor(q6qf$Income, levels=ordered)
```
```{r}
ggplot(q6qf, aes(y=happy, x=Income)) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle = -90))+ ylim(0, 3)+xlab("Income")+ylab("Happiness")+ggtitle("Income vs. Happiness")
```
```{r}
q6qf
```

The graph and data show that as a persons income increases the average happy rating increases. For the data, i used "Not too happy" = 1, "Pretty happy" = 2, "Very happy" = 3. This allowed me to run an average of each person for each income range. Although the happy slightly increase as income increases, we can see that they are all still right around a 2 average, which is pretty happy.

